---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
library(knitr)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

## Loading and preprocessing the data

1. Download the data if not exists and unzip the data.
```{r preprocessing}
library(knitr)
setwd("C:/Users/Sourav/Documents/Technical/Data Science/Coursera Data Science Specialization/Coursera_05_Reproducible Research/data")
if (!file.exists("./activity.zip"))
{
    fileurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(fileurl,destfile = "./activity.zip")
  
} 

if (!file.exists("./activity.csv"))
{
unzip("./activity.zip",exdir = getwd())
}
```
2. Read the data and Process/transform(removing NA rows) the data into a format suitable for your analysis.

```{r}
setwd("C:/Users/Sourav/Documents/Technical/Data Science/Coursera Data Science Specialization/Coursera_05_Reproducible Research/data")
actdat <- read.csv("activity.csv",header = TRUE)
actdat$date <- as.Date(actdat$date,"%Y-%m-%d")
actdatclean <- actdat[complete.cases(actdat[,1]),]
```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day and visulalize thorugh barplot.


```{r}
acttot <- aggregate(steps~date,data=actdatclean,sum)
library(ggplot2)
g <- ggplot(acttot,aes(date,steps))
g + geom_bar(stat = "identity") + geom_line(color="blue") + ggtitle("Total number of steps-day wise")
```

2. Make a histogram of the total number of steps taken each day
```{r}
g1 <- ggplot(acttot,aes(steps))
g1 <- g1 + geom_histogram() + labs(title="Histogram of the Total number of steps taken each day")
g1
```

3. The  mean and median of the total number of steps taken per day.
```{r}
mn <- mean(acttot$steps)
med <- median(acttot$steps)
```
The mean of the total number of steps is `r format(round(mn, 2), nsmall = 2)` and the median is `r med`.

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
dly_actp <- aggregate(steps~interval,data=actdatclean,mean)
g <- ggplot(dly_actp,aes(interval,steps))
g + geom_line(color="blue") + ggtitle("Time series plot of the 5-minute interval and the average number of steps taken")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
dly_actp[which.max(dly_actp$steps),][[1]]

```


## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
missing <- actdat[!complete.cases(actdat),]
dim(missing)[1] 
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

* Strategy is to replace with the mean for that 5-minute interval calculated above. So merging the full dataset NA with the interval average dataset.

```{r}
head(actdat)
head(dly_actp)
mrgfull <- merge(actdat,dly_actp,by="interval")
mrgfull$steps <- ifelse(is.na(mrgfull$steps.x),mrgfull$steps.y,mrgfull$steps.x)


```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
actnonadat <- mrgfull[order(mrgfull$date,mrgfull$interval),c("steps","date","interval")]
head(actnonadat)
```

4. Make a histogram of the total number of steps taken each day
```{r}
agractnonadat <- aggregate(steps~date,data=actnonadat,sum)
 g2 <- ggplot(agractnonadat,aes(steps))
 g2 <- g2 + geom_histogram() + ggtitle("The total number of steps taken each day on NA replaced data ")
 g2
 
```

5. Calculate and report the mean and median total number of steps 
```{r}
mn_new <- mean(agractnonadat$steps)
med_new <- median(agractnonadat$steps)
```

The mean of the NA replaced dataset is `r format(round(mn_new, 2), nsmall = 2)` and the median is `r format(round(med_new, 2), nsmall = 2)`.

6. Do these values differ from the estimates from the first part of the assignment?

* Earlier Mean was `r format(round(mn, 2), nsmall = 2)` and the new mean(after NA replacement) is `r format(round(mn_new, 2), nsmall = 2)`.

* Earlier Median was `r med` and the new mean(after NA replacement) is `r format(round(med_new, 2), nsmall = 2)`.

**Total number of steps comparison below.**
```{r}
library(gridExtra)
grid.arrange(g1, g2, ncol = 2)
```

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
actnonadat$day <- weekdays(actnonadat$date)
actnonadat$dayType <- ifelse(actnonadat$day %in% c("Monday","Tuesday","Wednesday","Thursday","Friday"),"Weekday","weekend")
actnonadat$dayType <- as.factor(actnonadat$dayType)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
actnonadat_intrvl <- aggregate(steps~interval+dayType,data=actnonadat,mean)

g <- ggplot(actnonadat_intrvl,aes(interval,steps))
g + geom_line() + facet_grid(dayType~.) + ggtitle("Avg steps at Intervals during weekdays and weekends")
```
